\documentclass[../sparc.tex]{subfiles}
\graphicspath{{\subfix{../images/}}}
\begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{I2C bus}
\label{section:i2c}
\index{Electronics!I2C bus}

\newglossaryentry{I2C}{
  name=I2C,
  description={Inter-Integrated Circuit}
}
\newglossaryentry{USB}{
  name=USB,
  description={Universal Serial Bus}
}
\newglossaryentry{PCI}{
  name=PCI,
  description={Peripheral component interconnect}
}
\newglossaryentry{I/O}{
  name=I/O,
  description={I/O -- ``Input/Output''}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{General information}

\textit{\gls{I2C}} also sometimes called $I^{2}C$ (should be read as
``I-squared-C'') is a data transfer bus that is used for connection between
integrated circuit inside electronic devices.  Let's discuss first what a ``data
bus'' is.  To put it simply a \textit{bus} in the context of the computer
architecture is some connection that is used for data transferring between
functional parts of a computer.

We can put data transfer buses roughly into two categories -- \textit{external}
buses and \textit{internal} buses.

External buses include \gls{USB} which is already familiar for many of us -- this
bus is used for connection of external devices (such as Arduino, a keyboard, a
computer mouse, a printer and so on) to a computer.  Such buses usually have
some standardized connector and fairly big length of the cable.

Internal buses include \gls{I2C} (already mentioned above) and \gls{PCI}/PCI
Express (used for example for connection of a network card to the computer
system board.)  Internal buses are usually designed for transferring data across
short distances, and don't allow hot-plugging of devices; thus some of the
internal buses don't even have a special connector and bus wires are soldered
down right to the boards.

The maximum length of connection between devices on \gls{I2C} bus should not
exceed 30 centimeters, otherwise data transfer will be unreliable.

\end{document}
