\documentclass[../sparc.tex]{subfiles}
\graphicspath{{\subfix{../images/}}}
\begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Arduino Platform}
\label{section:arduino}

\newglossaryentry{CNC}{name=CNC, description={Computer Numerical Control}}
\newglossaryentry{MCU}{name=MCU, description={Micro-Controller Unit}}

A micro-controller (or \gls{MCU}) is a programmable chip that consists of a
processor and a wide range of additional blocks.  MCUs are used to control
electronic devices: modern domestic devices, children toys, musical instruments
and industrial robots; upon MCUs are built 3D-printers and many other types of
\gls{CNC}.

One of the notable traits of an MCU is that it has high component integration
ratio: the space inside the chip is shared between a CPU, a RAM and non-volatile
memory.  The cost of MCUs may vary, as it heavily depends on the MCU features
and its purpose: the cheapest ones cost around few cents\cite{hackaday:padauk},
while on average consumer-grade MCUs have price tag starting from few dollars up
to few hundred dollars.

We will be working with Arduino MCU platform that provides convenient
programming interface.

Most of the Arduino platforms are built on AVR micro-controllers.

Here are some of the popular Arduino platforms:
\begin{itemize}
\item \textbf{Arduino Nano} -- One of the cheapest and most accessible Arduino
  versions. It has small size and modest number of input/output ports.
\item \textbf{Arduino Uno} -- A popular version of the platform that allows us to
  connect extension boards (or so called \emph{shields}) to it directly (we will
  talk about shields later.)
\item \textbf{Arduino Mega 2560} -- Big board that is based on the MCU ATmega
  2560.  The board has more than 50 digital input/output ports plus 16 analog
  ports.  Arduino Mega 2560 can be used as the basis for CNC (such as
  3D-printers.)
\item \textbf{Arduino Leonardo} -- This board is notable because it can
  ``pretend'' to be a USB-device -- or as programmers say, this board can
  \emph{emulate} it.
\end{itemize}

\section{Extension Boards (Shields)}
\label{subsection:arduino-shields}

Many Arduino versions allow us to connect to them not only standalone sensors
but complete extension boards (also known as \emph{shields}.)  Such shields are
connected to the top of an Arduino board.  Moreover often you can connect
another shield on top of the one already connected -- it's like a layered pancake
with Arduino on the bottom.  Shields provide different functions starting from
extending Arduino with Wi-Fi connectivity to the controlling full-fledged
\gls{CNC} (such as a 3D-printer or a laser cutter.)

\end{document}
